Cycle Route Uptake and Scenario Estimation (CRUSE): an approach for developing strategic cycle network planning tools

### 2024-03-25 paper.tex
Update tex

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Cycle Route Uptake and Scenario Estimation (CRUSE): an approach for developing strategic cycle network planning tools},
  pdfauthor={Robin Lovelace; Joey Talbot; Eugeni Vidal-Tortosa; Hussein Mahfouz; Elaine Brick; Peter Wright; Gary O'Toole; Dan Brennan; Suzanne Meade},
  pdfkeywords={Cycling, Open-source, Road Safety, Active Travel},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{keyword}
    Cycling \sep Open-source \sep Road Safety \sep 
    Active Travel
\end{keyword}
\end{frontmatter}

@@

Transport systems dominated by heavy and powerful private cars are
inefficient, dangerous, and unhealthy. Cars are a major and growing
source of greenhouse gas emissions causing climate change
\citep{winkler2023}, a leading cause of premature death and injury due
to road traffic collisions \citep{globals2018} , and a cause of disease
due to physical inactivity and air, noise and (from tyres) microplastic
pollution \citep{mattsson2023} \citep{welch2023} \citep{cavallaro2024}.
Transport is responsible for 23\% of global greenhouse gas (GHG)
emissions, 70\% of which are from road transport, with passenger cars
accounting for nearly half of transport emissions (around 10\% of global
emissions) \citep{jaramillo2022}. The transport system encourages,
enables and in some cases enforces unsustainable lifestyles. Services
that are only accessible by car lock-in car dependency
\citep{gray2001, shergold2012, motte-baumvol2010}.

@@

Building cycle networks represents a relatively `quick win' within the
context of decarbonization \citep{brand2020} and sustainable mobility
\citep{burns2013}. Although cycling uptake appears on the surface to
only relate to the `shift' part of the ASI framework, closer
consideration of the knock-on impacts of cycling uptake shows that it
can also help avoid unnecessary trips \citep{nello-deakin2020}.
Furthermore, highly efficient ebikes --- which are seeing rapid uptake
in Ireland and other countries --- outperform electric cars, which are
too heavy and expensive. Over-reliance on electric cars could slow the
transition away from car dependency and inadvertently enable ``high
travel lock-in'' \citep{anable2019}. At the European level, the European
Union has a target of reducing GHG emissions by 55\% by 2030, compared
to 1990 levels, and to achieve `net-zero' by 2050 \citep{rosenow2022}.
Climate change mitigation is a major motivation for cycle network plans
\citep{scappini2022}.

@@

At the regional level within Ireland, the recently published Greater
Dublin Area Transport Strategy \citep{greater} reveals the high support
for and capability for cycling: ``nearly a quarter of adults cycle at
least once a week in the Dublin Metropolitan Area'' with cycling in the
Dublin area taking up to 60,000 cars off the road today. Extrapolating
this on a per population basis across Ireland, with around 40\% of the
population living in Dublin, suggests that around 150,000 cars could be
removed nationwide just by achieving Dublin levels of cycling in all
counties (notwithstanding existing cycling trips and differences in trip
distances). 7 in 10 trips in Ireland are by car \citep{national} .
Cycling has the potential to replace a large proportion of these trips:
``A high priority must also be given to cyclists, because trips by this
mode have the potential to replace trips by private car, most
specifically for short to medium distance trips, but increasingly for
longer trips as ebikes extend the range of this mode'' \citep{greater}.

@@

This paper describes an approach to strategic cycle network planning
tool development that is open, scalable, and evidence-based. We label
the approach Cycle Route Uptake and Scenario Estimation (CRUSE) and
present a case study of its application in Ireland, the results of which
are publicly available at \href{https://cruse.bike/}{cruse.bike} in the
CRUSE Tool for Ireland, the landing page of which is illustrated in
Figure~\ref{fig-landing}. In Section~\ref{sec-methods}, we outline the
methods used to generate the evidence presented in the CRUSE Tool. In
Section~\ref{sec-results}, we present the results of the CRUSE Tool,
including estimates of current cycling levels and future cycling
potential at the national, regional and local levels. In
Section~\ref{sec-discussion}, we discuss limitations and possible future
improvements to the approach, and the implications of the results for
Ireland before making concluding remarks in
Section~\ref{sec-conclusions}.

\section{Tools for estimating cycling
potential}\label{tools-for-estimating-cycling-potential}

@@

A feature of active travel interventions is that they require dense
networks of routes to be effective \citep{parkin2018}. This means that
high levels of geographic resolution are needed in the data used to
estimate cycling potential. However, datasets on travel patterns are
often only available at the level of administrative zones. The Central
Statistics Office (CSO) in Ireland provides Place of Work, School or
College Census of Anonymized Records
(\href{https://www.cso.ie/en/census/census2016reports/powscar/}{POWSCAR})
data on the number of people traveling to work and school at the
Electoral Division (ED) level, for example.

In the PCT, the method used to convert OD data to route networks was to
calculate a single route between the population weighted centroids of
the zones associated with each OD pair. This method works fine when the
OD data represents movement between small areas, but was not appropriate
for generating route networks from the POWSCAR data because zone
centroids are so far apart that the resulting route networks would be
sparse and unrealistic. To tackle this issue we developed a new method
for OD data disaggregation called `jittering' \citep{lovelace2022}. The
method works by first disaggregating the OD data based on a
`disaggregation threshold' and then assigning each disaggregated
`sub-OD' pair to `subpoints' within each zone. As shown in
Figure~\ref{fig-dublin}, the resulting route networks are dense, even in
rural areas.

@@

However, travel to school and work only constitute around only 30\% of
all trips in Ireland. To overcome this issue, we developed a spatial
interaction modeling methodology to estimate the number of trips between
each OD pair for additional trip purposes. The classification of trip
purposes used in the CRUSE Tool was guided mainly by the trip purpose
classification found within the National Household Travel Survey (NHTS),
but with the addition of categories based on the comprehensive POWSCAR
data, and the need to include recreational trips and multi-stage trips
(not yet implemented). An overview of the trip purposes used in CRUSE is
presented in Figure~\ref{fig-trip-purposes}.

@@

We calculated three network types for each disaggregated (`jittered') OD
pair: `Fastest', `Balanced' and `Quietest'. As outlined on the
CycleStreets'
\href{https://www.cyclestreets.net/help/journey/howitworks/}{website},
the `Fastest' routes minimize journey time, accounting for traffic
lights and surface type. The `Quietest' route type minimizes busy
sections of road, allowing high `diversion factors' from the fastest
route on quieter but less direct ways (often avoiding roads and
interactions with motor vehicles altogether where possible). The
`Balanced' route type is a compromize between the two, minimising
journey time while avoiding the busiest roads. Allowing users to switch
between these three route types enables them to consider the trade-offs
between directness and `cycle friendliness' when planning new
infrastructure, as shown in Figure~\ref{fig-route-types}.

@@

These statistics can be revealing: in Kildare it suggests that around
half of all cycling activity occurs on network segments that are hostile
(with a high inferred level of traffic stress), even when efforts are
taken to avoid busy segments. For the Fastest network, which may be more
realistic for utility cycling, the proportion of cycling on hostile
segments is even higher at 75\%.

@@

\caption{\label{fig-route-types}Illustration of the Quietest (left) and
Fastest (right) route networks for Naas (top) and Newbridge (bottom).}

@@

The Climate Action Plan scenario models the transport emissions
reductions targeted in the Irish Government's
\href{https://www.gov.ie/en/publication/6223e-climate-action-plan-2021/}{Climate
Action Plan 2021}, which aims for a 51\% cut in overall greenhouse gas
emissions by 2030 as part of the pathway to net-zero emissions by 2050.
For transport, this includes 500,000 extra walking, cycling and public
transport trips per day by 2030. In terms of car travel, the target is
to ``Increas{[}e{]} the proportion of kilometers driven by passenger
electric cars to between 40 and 45\% by 2030, in addition to a reduction
of 10\% in kilometers driven by the remaining internal combustion engine
cars.'' This equates to a 5.5 to 6\% reduction in total car km driven.

To model this decrease in car driver km, cycling uptake increases in
line with the Go Dutch scenario. However, we only model shift from
driving to cycling. There is no shift from other modes of transport to
cycling.

@@

Under the Go Dutch scenario cycling reaches levels equivalent to those
found in the Netherlands, taking account of the effects of route
hilliness (measured as mean gradient) and route distance. This scenario
uses the same model that was used PCT, allowing trips to shift from any
other mode to cycling \citep{lovelace2017}.

@@

The CRUSE web application is statically hosted, meaning that it can be
accessed without the need for a server running software such as the R
package `shiny' or Python packages such as `streamlit' or `flask' in the
background \citep{wickham2021}. Typical intended user stories are
illustrated in Figure~\ref{fig-user-stories}, which shows that the tool
is designed to be used by both professional and non-professional users.
For the main target audience, professional transport planners working at
the county level, the tool provides a range of outputs, including
estimates of cycling potential at the county and network level. The
provision of balanced (the default), quietest and fastest route networks
enables planners to consider the trade-offs between directness and
`cycle friendliness' when planning new infrastructure. The tool also
provides data downloads, enabling estimates of cycling potential on the
network to be visualised and analysed with other tools such as QGIS,
Python or R.

@@

Aside from the widely varying sizes and route networks of each city,
with Dublin having by far the densest network, it is clear from the
visualisation of cycling friendliness that there are many gaps in the
networks. Dublin has had received most investment in cycling
infrastructure and this is apparent from the high proportion of the
route network that is green, even in the `Fastest' route network
representing routes that prioritize speed and directness over comfort
and safety. Dublin also has the highest mode share of cycling of the
four cities, adding to the evidence base showing that investment in
cycling infrastructure is effective in increasing cycling uptake.

Although Dublin City has made progress, there are still many parts of
the Fastest route network, and even some parts of the Balanced and
Quietest route networks, that are not cycle friendly and which have high
cycling potential. According to a recent report, 71\% of residents in
the Dublin Metropolitan Area support ``more cycle tracks along roads,
physically separated from traffic and pedestrians'' \citep{walking2021}.
The results for the Dublin area could help prioritise investment in
cycling infrastructure in the city.

@@

The aim of the paper was to describe the design, features and potential
use of the `CRUSE' approach and resulting tool. As outlined in the
previous section, the CRUSE Tool provides evidence on current and
potential future cycling levels across Ireland down to the street level,
with potential assigned to `fastest', `balanced' and `quietest' route
networks. This provision of multiple scenarios of behavior change
\emph{and} multiple scenarios of investment, for example in cycling
infrastructure next to major roads vs quiet residential streets, is a
key feature of the CRUSE Tool.

@@

As the benefits of strategic planning for cycling become more apparent
\citep{scappini2022} we expect the demand for tools such as CRUSE to
grow. However, the approach is not without limitations and does not meet
all requirements in Ireland or indeed any country where the methods are
applied.

A limitation of the approach as implemented in the CRUSE tool is that
the results are `static', reducing its ability to support monitoring of
growth in cycling, upgrades of existing cycling infrastructure and
dynamic (year-by-year) exposure information to estimate crash rates.
Exposure information is a key part of the the European RISM Directive,
for reporting collision rates of vulnerable road users by 2024. These
factors mean that the tool can be seen as an open access `leverage
point', providing key information for and supporting ambitious plans in
many aspects the planning system, from network design to post-build
monitoring \citep{lovelace2020}.

>>>>

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Cycle Route Uptake and Scenario Estimation (CRUSE): an approach for developing strategic cycle network planning tools},
  pdfauthor={Robin Lovelace; Joey Talbot; Eugeni Vidal-Tortosa; Hussein Mahfouz; Elaine Brick; Peter Wright; Gary O'Toole; Dan Brennan; Suzanne Meade},
  pdfkeywords={Cycling, Open-source, Road Safety, Active
Travel, Transport Planning, Collaborative Planning},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{keyword}
    Cycling \sep Open-source \sep Road Safety \sep Active
Travel \sep Transport Planning \sep 
    Collaborative Planning
\end{keyword}
\end{frontmatter}

@@

\subsection{Background}\label{background}

Transport systems dominated by heavy and powerful private cars are
inefficient, dangerous, and unhealthy. Cars are a major and growing
source of greenhouse gas (GHG) emissions \citep{winkler2023}, a leading
cause of premature death and injury due to road traffic collisions
\citep{globals2018}, and a cause of disease due to physical inactivity
and air, noise and microplastic pollution \citep{mattsson2023}
\citep{welch2023} \citep{cavallaro2024}. Transport is responsible for
23\% of GHG emissions, 70\% of which are from road transport, with
passenger cars accounting for nearly half of transport emissions (around
10\% of global emissions) \citep{jaramillo2022}. The transport system
encourages, enables and in some cases enforces unsustainable lifestyles.
Services that are only accessible by car lock-in car dependency
\citep{gray2001, shergold2012, motte-baumvol2010}.

@@

Building cycle networks represents a relatively `quick win' within the
context of decarbonization \citep{brand2020} and sustainable mobility
\citep{burns2013}. Although cycling uptake appears on the surface to
only relate to the `shift' part of the ASI framework, closer
consideration of the knock-on impacts of cycling uptake shows that it
can also help avoid unnecessary trips \citep{nello-deakin2020}.
Furthermore, highly efficient ebikes --- which are seeing rapid uptake
--- outperform electric cars, which are too heavy and expensive, for the
majority of trips. Over-reliance on electric cars could slow the
transition away from car dependency and inadvertently enable ``high
travel lock-in'' \citep{anable2019}. At the European level, the European
Union has a target of reducing GHG emissions by 55\% by 2030, compared
to 1990 levels, and to achieve `net-zero' by 2050 \citep{rosenow2022}.
Climate change mitigation is a major motivation for cycle network plans
\citep{scappini2022}.

@@

At the regional level within Ireland, the recently published Greater
Dublin Area Transport Strategy \citep{greater} reveals the high support
for and capability for cycling: ``nearly a quarter of adults cycle at
least once a week in the Dublin Metropolitan Area'' with cycling in the
Dublin area taking up to 60,000 cars off the road today. Extrapolating
this on a per population basis across Ireland, with around 40\% of the
population living in Dublin, suggests that around 150,000 cars could be
removed nationwide just by achieving Dublin levels of cycling in all
counties (notwithstanding existing cycling trips and differences in trip
distances). Seven in ten trips in Ireland are by car \citep{national}.
Cycling has the potential to replace a large proportion of these trips:
``A high priority must also be given to cyclists, because trips by this
mode have the potential to replace trips by private car, most
specifically for short to medium distance trips, but increasingly for
longer trips as ebikes extend the range of this mode'' \citep{greater}.

@@

\subsection{Aim and content}\label{aim-and-content}

This paper describes an approach to strategic cycle network planning
tool development that is open, scalable, and evidence-based. We label
the approach Cycle Route Uptake and Scenario Estimation (CRUSE) and
present a case study of its application in Ireland, the results of which
are publicly available at \href{https://cruse.bike/}{cruse.bike}.

In Section~\ref{sec-tools}, we review existing current tools for
estimating cycling potential. In Section~\ref{sec-methods}, we outline
the methods used to generate the evidence presented in the CRUSE Tool.
In Section~\ref{sec-results}, we present the results of the CRUSE Tool,
including estimates of current cycling levels and future cycling
potential at the national, regional and local levels. In
Section~\ref{sec-discussion}, we discuss limitations and possible future
improvements to the approach, and the implications of the results for
Ireland before making concluding remarks in
Section~\ref{sec-conclusions}.

\section{Tools for estimating cycling potential}\label{sec-tools}

@@

A feature of active travel interventions is that they require dense
networks of routes to be effective \citep{parkin2018}. This means that
data with high levels of geographic resolution are necessary to estimate
cyclical potential. However, datasets on travel patterns are often only
available at the level of administrative zones. The Central Statistics
Office (CSO) in Ireland provides Place of Work, School or College Census
of Anonymized Records
(\href{https://www.cso.ie/en/census/census2016reports/powscar/}{POWSCAR})
data on the number of people traveling to work and school at the
Electoral Division (ED) level, for example.

In the PCT, the method used to convert OD data to route networks was to
calculate a single route between the population weighted centroids of
the zones associated with each OD pair. This method works fine when the
OD data represents movement between small areas, but was not appropriate
for generating route networks from the POWSCAR data because zone
centroids are so far apart that the resulting route networks would be
sparse and unrealistic. To tackle this issue we developed a new method
for OD data disaggregation called `jittering'. The method works by first
disaggregating the OD data based on a `disaggregation threshold' and
then assigning each disaggregated `sub-OD' pair to `subpoints' within
each zone. As shown in Figure~\ref{fig-dublin}, the resulting route
networks are dense, even in rural areas.

@@

However, travel to school and work only constitute around only 30\% of
all trips in Ireland. To overcome this issue, we developed a spatial
interaction modeling methodology to estimate the number of trips between
each OD pair for additional trip purposes. The classification of trip
purposes used in the CRUSE Tool was guided mainly by the trip purpose
classification found within the National Household Travel Survey (NHTS),
but with the addition of categories based on the comprehensive POWSCAR
data, and the need to include recreational trips and multi-stage trips.
An overview of the trip purposes used in CRUSE is presented in
Figure~\ref{fig-trip-purposes}.

@@

We calculated three network types for each disaggregated (`jittered') OD
pair: `Fastest', `Balanced' and `Quietest'. As outlined on the
CycleStreets'
\href{https://www.cyclestreets.net/help/journey/howitworks/}{website},
the fastest routes minimize journey time, accounting for traffic lights
and surface type. The `Quietest' route type minimizes busy sections of
road, allowing high `diversion factors' from the fastest route on
quieter but less direct ways (often avoiding roads and interactions with
motor vehicles altogether where possible). The `Balanced' route type is
a compromize between the two, minimising journey time while avoiding the
busiest roads. Allowing users to switch between these three route types
enables them to consider the trade-offs between directness and `cycle
friendliness' when planning new infrastructure, as shown in
Figure~\ref{fig-route-types}.

@@

These statistics can be revealing: in Kildare it suggests that around
half of all cycling activity occurs on network segments that are hostile
(with a high inferred level of traffic stress), even when efforts are
taken to avoid busy segments. For the fastest network, which may be more
realistic for utility cycling, the proportion of cycling on hostile
segments is even higher at 75\%.

@@

\caption{\label{fig-route-types}Illustration of the quietest (left) and
fastest (right) route networks for Naas (top) and Newbridge (bottom) in
County Kildare.}

@@

The Climate Action Plan scenario models the transport emissions
reductions targeted in the Irish Government's
\href{https://www.gov.ie/en/publication/6223e-climate-action-plan-2021/}{Climate
Action Plan 2021}, which aims for a 51\% cut in overall GHG emissions by
2030 as part of the pathway to net-zero emissions by 2050. For
transport, this includes 500,000 extra walking, cycling and public
transport trips per day by 2030. In terms of car travel, the target is
to ``Increas{[}e{]} the proportion of kilometers driven by passenger
electric cars to between 40 and 45\% by 2030, in addition to a reduction
of 10\% in kilometers driven by the remaining internal combustion engine
cars.'' This equates to a 5.5 to 6\% reduction in total car km driven.

To model this decrease in car km driven, cycling uptake increases in
line with the Go Dutch scenario. However, we only model shift from
driving to cycling. There is no shift from other modes of transport to
cycling.

@@

Under the Go Dutch scenario cycling reaches levels equivalent to those
found in the Netherlands, taking account of the effects of route
hilliness (measured as mean gradient) and route distance. This scenario
uses the same model as the PCT, allowing trips to shift from any other
mode to cycling \citep{lovelace2017}.

@@

The CRUSE web application is statically hosted, meaning that it can be
accessed without the need for a server running software such as the R
package `shiny' or Python packages such as `streamlit' or `flask' in the
background \citep{wickham2021}. Its landing page is illustrated in
Figure~\ref{fig-landing}).

Typical intended user stories are illustrated in
Figure~\ref{fig-user-stories}, which shows that the tool is designed to
be used by both professional and non-professional users. For the main
target audience, professional transport planners working at the county
level, the tool provides a range of outputs, including estimates of
cycling potential at the county and network level. The provision of
balanced (the default), quietest and fastest route networks enables
planners to consider the trade-offs between directness and `cycle
friendliness' when planning new infrastructure. The tool also provides
data downloads, enabling estimates of cycling potential on the network
to be visualised and analysed with other tools such as QGIS, Python or
R.

@@

Aside from the widely varying sizes and route networks of each city,
with Dublin having by far the densest network, it is clear from the
visualisation of cycling friendliness that there are many gaps in the
networks. Dublin has had received most investment in cycling
infrastructure and this is apparent from the high proportion of the
route network that is green, even in the fastest route network
representing routes that prioritize speed and directness over comfort
and safety. Dublin also has the highest mode share of cycling of the
four cities, adding to the evidence base showing that investment in
cycling infrastructure is effective in increasing cycling uptake.

Although Dublin City has made progress, there are still many parts of
the fastest route network, and even some parts of the Balanced and
Quietest route networks, that are not cycle friendly and which have high
cycling potential. According to a recent report, 71\% of residents in
the Dublin Metropolitan Area support ``more cycle tracks along roads,
physically separated from traffic and pedestrians'' \citep{walking2021}.
The results for the Dublin area could help prioritise investment in
cycling infrastructure in the city.

@@

The aim of the paper was to describe the design, features and potential
use of the `CRUSE' approach and resulting tool. As outlined in the
previous section, the CRUSE Tool provides evidence on current and
potential future cycling levels across Ireland down to the street level,
with potential assigned to fastest, balanced and quietest route
networks. This provision of multiple scenarios of behavior change
\emph{and} multiple scenarios of investment, for example in cycling
infrastructure next to major roads vs quiet residential streets, is a
key feature of the CRUSE Tool.

@@

As the benefits of strategic planning for cycling become more apparent
\citep{scappini2022}, we expect the demand for tools such as CRUSE to
grow. However, the approach is not without limitations and does not meet
all requirements in Ireland or indeed any country where the methods are
applied.

A limitation of the approach as implemented in the CRUSE Tool is that
the results are `static', reducing its ability to support monitoring of
growth in cycling, upgrades of existing cycling infrastructure and
dynamic (year-by-year) exposure information to estimate crash rates.
Exposure information is a key part of the the European RISM Directive,
for reporting collision rates of vulnerable road users by 2024. These
factors mean that the tool can be seen as an open access `leverage
point', providing key information for and supporting ambitious plans in
many aspects the planning system, from network design to post-build
monitoring \citep{lovelace2020}.
